Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-07-22 threshold/draft.Rnw
Fixed more typos and figure label

The following lemma expresses the Euclidean distance between any two columns of the
matrix $A$ as a weighted sum of projected interval distances along the SVD
basis vectors forming the columns of $V$.
\begin{lemma}
Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix
$A$. Then
\begin{equation}\label{vdist}
\|a_i - a_j\|^2 =
\sigma_1^2 |v_{1i} - v_{1j}|^2 + 
\sigma_2^2 |v_{2i} - v_{2j}|^2 + \cdots + 
\sigma_k^2 |v_{ki} - v_{kj}|^2,
\end{equation}
where $v_{kj}$ denotes the $jth$ position in the vector $v_k$ (that is, the
$[k,j]$ entry in the matrix $V$).
\end{lemma}

@@

In practice, many pairs of vectors may be ruled out in this fashion by
inspecting distances in low-dimensional subspaces, substantially reducing the
complexity of computing thresholded correlation matrices.  Approaches that
cheaply omit pairs of vectors from consideration are called \emph{pruning
methods}, and a variety pruning methods have been described elsewhere for the
thresholded correlation problem; see, for example\cite{zf,mnl} and the
references therein.  The approach described here prunes along an intuitive path
of decreasing variance defined by the data and computed by the SVD.

@@

\begin{figure}[!ht]
\label{fig1}
\begin{center}
<<eval=TRUE, echo=FALSE, fig=TRUE, height=5, width=7>>=
library(biclust)
data(BicatYeast)
A=t(BicatYeast)
s=svd(A)
i=order(s$v[,1])
plot(s$v[i,1], ylab=expression(P * v[1]), xlab="Sorted index", pch='.')
t=0.99
abline(h=c(-0.05,-0.05+sqrt(2*(1-t))/s$d[1]),col=2,lwd=0.5)
@
\caption{Example plot of the ordered entries of $v_1$ for the BicatYeast example from Section \ref{examples}. The red lines illustrate the interval corresponding threshold $t=0.99$. Points farther apart than this interval correspond to column vectors that do not meet the correlation threshold.}
\end{center}
\end{figure}

Let $D^i\in\mathcal{R}^{n-i, n}$, $i=1,2,\ldots,n-1$, be the $i$th order finite
difference matrix with $-1$ along the diagonal, $1$ along the $i$th
super-diagonal, and zeros elsewhere.  Consider just adjacent points of $P v_1$.
Then $D^1P v_1$ consists of the differences of adjacent entries in $P v_1$.
Entries such that $\sigma_1^2 (D^1 P v_1)^2 > 2(1-t)$
correspond to pairs of vectors that do not meet the correlation threshold,
(exponentiation is applied to the vector element-wise).
However, this observation is unlikely to rule out many pairs of vectors in
most problems. For example, the minimum distance between the points shown in
the example in Figure \ref{fig1} is less that $\sqrt{2(1-t)}/\sigma$, and no
pruning at all occurs.

Including more terms from Equation \ref{vdist} increases our ability to prune
pairs of vectors below the correlation threshold. For example, including
only one more term finds that 72\% of adjacent vectors fall below the
correlation threshold
in the example shown in Figure \ref{fig1} and are pruned by identifying
indices such that
\[
\sigma_1^2 (D^1 P v_1)^2 +
\sigma_2^2 (D^1 P v_2)^2 > 2(1-t),
\]
where exponentiation is applied element-wise to the vectors.
Note that we use the permutation $P$ defined by the order of vector $v_1$
throughout. Including five terms prunes over 97\% of adjacent vector pairs,
leaving only 11 candidate pairs of adjacent vectors.
In general, using $p\le k$ terms to prune pairs of vectors below the
correlation threshold boils down to evaluating the expression
\begin{equation}\label{proj}
(D^1 P V_{1:p} )^2 s_{1:p}^2 > 2(1-t),
\end{equation}
where 
$V_{1:p}$ denotes the first $p$ columns of the
matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values,
and exponentiation is applied to vectors element-wise.

>>>>

The following lemma expresses the Euclidean distance between any two columns of the
matrix $A$ as a weighted sum of projected interval distances along the SVD
basis vectors forming the columns of $V$.
\begin{lemma}
Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix
$A$. Then
\begin{equation}\label{vdist}
\|a_i - a_j\|^2 =
\sigma_1^2 |v_{1i} - v_{1j}|^2 + 
\sigma_2^2 |v_{2i} - v_{2j}|^2 + \cdots + 
\sigma_k^2 |v_{ki} - v_{kj}|^2,
\end{equation}
where $v_{kj}$ denotes the $jth$ position in the vector $v_k$ (that is, the
$[j,k]$ entry in the matrix $V$).
\end{lemma}

@@

In practice, many pairs of vectors may be ruled out in this fashion by
inspecting distances in low-dimensional subspaces, substantially reducing the
complexity of computing thresholded correlation matrices.  Approaches that
cheaply omit pairs of vectors from consideration are called \emph{pruning
methods}, and a variety of pruning methods have been described elsewhere for
the thresholded correlation problem; see, for example\cite{zf,mnl} and the
references therein.  The approach described here prunes along an intuitive path
of decreasing variance defined by the data and computed by the SVD.

@@

\begin{figure}[!ht]
\begin{center}
<<eval=TRUE, echo=FALSE, fig=TRUE, height=5, width=7>>=
library(biclust)
data(BicatYeast)
A=t(BicatYeast)
s=svd(A)
i=order(s$v[,1])
plot(s$v[i,1], ylab=expression(P * v[1]), xlab="Sorted index", pch='.')
t=0.99
abline(h=c(-0.05,-0.05+sqrt(2*(1-t))/s$d[1]),col=2,lwd=0.5)
@
\caption{
Example plot of the ordered entries of $v_1$ for the BicatYeast example from Section \ref{examples}. The red lines illustrate the interval corresponding threshold $t=0.99$. Points farther apart than this interval correspond to column vectors that do not meet the correlation threshold.
\label{fig1}
}
\end{center}
\end{figure}

Let $D^i\in\mathcal{R}^{n-i, n}$, $i=1,2,\ldots,n-1$, be the $i$th order finite
difference matrix with $-1$ along the diagonal, $1$ along the $i$th
super-diagonal, and zeros elsewhere.  Consider just adjacent points of $P v_1$.
Then $D^1P v_1$ consists of the differences of adjacent entries in $P v_1$.
Entries such that $\sigma_1^2 (D^1 P v_1)^2 > 2(1-t)$ correspond to pairs of
vectors that do not meet the correlation threshold, where here and below
exponentiation applied to a vector denotes element-wise exponentiation.
However, this observation is unlikely to rule out many pairs of vectors in most
problems. For example, the minimum distance between the points shown in the
example in Figure \ref{fig1} is less that $\sqrt{2(1-t)}/\sigma_1$, and no
pruning at all occurs.

Including more terms from Equation \ref{vdist} increases our ability to prune
pairs of vectors below the correlation threshold. For example, including
only one more term finds that 72\% of adjacent vectors fall below the
correlation threshold
in the example shown in Figure \ref{fig1} and are pruned by identifying
indices such that
\[
\sigma_1^2 (D^1 P v_1)^2 +
\sigma_2^2 (D^1 P v_2)^2 > 2(1-t).
\]
Note that we use the permutation $P$ defined by the order of vector $v_1$
throughout. Including five terms prunes over 97\% of adjacent vector pairs,
leaving only 11 candidate pairs of adjacent vectors.
In general, using $p\le k$ terms to prune pairs of vectors below the
correlation threshold boils down to evaluating the expression
\begin{equation}\label{proj}
(D^1 P V_{1:p} )^2 s_{1:p}^2 > 2(1-t),
\end{equation}
where 
$V_{1:p}$ denotes the first $p$ columns of the
matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values.
